---
title: "Data cleaning"
author: "Emma Ning"
date: "June 11^th^, 2020"
output:
  pdf_document:
    fig_caption: yes
    fig_height: 4
    fig_width: 5
    highlight: tango
  word_document:
    toc: yes
    toc_depth: '2'
  html_document:
    df_print: kable
    highlight: tango
    theme: sandstone
    toc: yes
    toc_depth: 2
    toc_float:
      collapsed: yes
---

```{r message = FALSE, warning = FALSE}
library(readxl)
#library(lubridate)
library(tidyverse)
library(psych) # Cronbach's alpha and summary stats etc.
library(plotrix) # for se calculation

setwd("/Volumes/GoogleDrive/My Drive/Summer2020/IndianaUniversity/June")

nifd <- read_excel("NIFD_version3.xlsx")
```



### Preliminary data visualizations + Data cleaning
```{r warning = FALSE}
# filtering out variables we don't need
nifd <- nifd %>% 
  select(-SITE)

# GENDER descriptive stats
nifd <- nifd %>% 
  mutate(gender.f = as.factor(GENDER)) # convert numeric gender to factor
gender <- nifd %>% 
  distinct(LONI_ID, gender.f)
table(gender$gender.f)

# labelling education variable
nifd$edu_level[nifd$EDUCATION <= 12 ] <- "HS/HS grad"
nifd$edu_level[nifd$EDUCATION > 12 & nifd$EDUCATION <= 16] <- "College/College grad"
nifd$edu_level[nifd$EDUCATION > 16] <- "post College"
# convert edu_level to factor for summarizing
nifd <- nifd %>% 
  mutate(edu_level.f = as.factor(edu_level))
#summary(nifd$edu_level.f)
ggplot(nifd, aes(x = edu_level.f)) + 
  geom_bar() + geom_text(stat='count', aes(label = ..count..), vjust = -0.2) # bar plot for education
#ggsave("education.png", height = 7, width = 12)

# WRANGLING DATES
# age at the start of data recording
# creating decimal dob for participants
nifd$dob_decimal <- lubridate::decimal_date(nifd$DOB)
nifd$appt_date_decimal <- lubridate::decimal_date(nifd$CLINICAL_LINKDATE)
nifd <- nifd %>% 
  mutate(age_at_first_appt = appt_date_decimal - dob_decimal)

# create a separate dataset for participants ID, their first visit, and their age
first_visit <- nifd %>% 
  select(LONI_ID, DOB, CLINICAL_LINKDATE, age_at_first_appt)
first_visit <- first_visit %>% 
  group_by(LONI_ID) %>% 
  filter(row_number()==1) # select the first row for each unique participant ID (their 1st visit)

# check for missing according to ID
#check1 <- subset(nifd, LONI_ID == "1_S_0349") # this person does not have anything
#check2 <- subset(first_visit, age_at_first_appt < 45) # check for people aged under 45 at first visit
#check3 <- subset(nifd, LONI_ID == "1_S_0143") # this 36 year old person has data, keep

# get rid of person aged 7 from dataset
#nifd <- subset(nifd, LONI_ID!= "1_S_0349")
#first_visit <- subset(first_visit, LONI_ID!= "1_S_0349")
# visualize age distributions at first visit
ggplot(first_visit, aes(x = age_at_first_appt)) + geom_histogram(binwidth = 1) + 
  ggtitle("Age distribution of participants at first visit") 

# calculate time lapse between each visit for each participant
visits_time_interval <- nifd %>% # and saves this info into a new dataframe
  select(LONI_ID, VISIT_NUMBER, CLINICAL_LINKDATE, appt_date_decimal) 

  #spread(VISIT_NUMBER, appt_date_decimal, sep = "")

# sort the visits_time_interval by subject ID and visit number, in ascending order
visits_time_interval <- visits_time_interval[with(visits_time_interval, order(LONI_ID, VISIT_NUMBER)), ]
# create new column in dataframe that calculates time interval difference
visits_time_interval <- visits_time_interval %>% 
  group_by(LONI_ID) %>% 
  mutate(visits_time_diff = appt_date_decimal - lag(appt_date_decimal, default = appt_date_decimal[1]))


# plotting visit times
nifd <- nifd %>% 
  mutate(visit_number.f = as.factor(VISIT_NUMBER))
ggplot(nifd, aes(x = visit_number.f)) + geom_bar() + 
  geom_text(stat='count', aes(label = ..count..), vjust = -0.2) +
  ggtitle("Summary of visit numbers")
#ggsave("numVisits.png", height = 7, width = 12)

# calculate the number of MMSE visits distribution for the sample
nifd <- nifd %>% 
  mutate(MMSE_YN = if_else(MMSE_TOT %in% NA, 0, 1)) %>% # whether they have a MMSE visit or not (YN)
  mutate(CDR_YN = if_else(CDR_TOT %in% NA, 0, 1))
# new df called visits
visits <- nifd %>% 
  select(LONI_ID, MMSE_TOT, MMSE_YN, CDR_TOT, CDR_YN) %>% 
  group_by(LONI_ID) %>% 
  mutate(mmse_total_visits = sum(MMSE_YN), cdr_total_visits = sum(CDR_YN))

visits <- visits %>% 
  group_by(LONI_ID) %>% 
  filter(row_number()==1) %>% 
  select(-MMSE_YN, -MMSE_TOT, -CDR_YN, -CDR_TOT)

ggplot(visits, aes(x = as.factor(mmse_total_visits)))+ geom_bar() + 
  geom_text(stat='count', aes(label = ..count..), vjust = -0.2) +
  ggtitle("Total MMSE visits")
#ggsave("MMSEvisits.png", height = 7, width = 12)
ggplot(visits, aes(x = as.factor(cdr_total_visits)))+ geom_bar() + 
  geom_text(stat='count', aes(label = ..count..), vjust = -0.2) + 
  ggtitle("Total CDR visits")
#ggsave("CDRvisits.png", height = 7, width = 12)
```


## Visualizing changes over time
### For semantic & non-fluent groups over time
Language behavioral measures:
BNT (Boston Naming Test) - BNTCORR
Phonemic Fluency - DCORR
Semantic Fluency - ANCORR
Peabody Picture Vocab 
Pyramids & Palm Trees - PPTP_COR
California Verbal Learning Test 

Use MMSE as baseline measure.
```{r}
lang_behav_measures <- nifd %>% 
  select(LONI_ID, DX, VISIT_NUMBER, CLINICAL_LINKDATE, dob_decimal, age_at_first_appt, gender.f, edu_level.f,
         MMSE_DCDATE, MMSE_TOT, DCORR, ANCORR, BNTCORR, PPTP_DCDATE, PPTP_COR, 
         TRCOTOT:RECOG, PPVTVRB:PPVTINA)
lang_behav_measures$clinical_date_decimal <- lubridate::decimal_date(lang_behav_measures$CLINICAL_LINKDATE)
# merge this dataset with the time interval between visits dataset
lang_behav_measures <- merge(lang_behav_measures, visits_time_interval, by = c("LONI_ID", "CLINICAL_LINKDATE"))
# get rid of redundant columns & change names
lang_behav_measures <- lang_behav_measures %>% 
  select(-VISIT_NUMBER.x) %>% 
  rename(visit_number = VISIT_NUMBER.y) %>% 
  mutate(diagnosis.f = as.factor(DX)) %>% 
  select(LONI_ID, DX, diagnosis.f, CLINICAL_LINKDATE, dob_decimal, age_at_first_appt, clinical_date_decimal:visits_time_diff,
         gender.f:PPVTINA) # reorder variable display order

# create another dataset with diagnosis, lang behav measures, and their time lapse between visits
# average values according to different diagnoses (CON, PNFA, SV)

# first calculate how many distinct diagnoses there are
# diagnosis_counts <- lang_behav_measures %>% 
#   distinct(LONI_ID, diagnosis.f)
# table(diagnosis_counts$diagnosis.f)

# CON, PNFA, SV counts in order: 126, 34, 37
# then calculate 
temp1 <- lang_behav_measures %>% 
  select(LONI_ID, diagnosis.f, visit_number, visits_time_diff, DCORR) %>% 
  mutate(time_diff_round = plyr::round_any(visits_time_diff, 0.5)) %>% # rounding in increments of .5
  filter(!is.na(DCORR)) # get rid of NA values for col DCORR

# if we were to use the rounded time difference
# calculate sum of DCORR score based on groups of diagnoses type and time differences round
# temp2 <- temp1 %>% 
#   group_by(diagnosis.f, time_diff_round) %>% 
#   mutate(sum_DCORR = sum(DCORR, na.rm = TRUE))

# if we were to use visit numbers
# calculate sum of DCORR score based on groups of diagnoses type and visit number
temp2 <- temp1 %>% 
  group_by(diagnosis.f, visit_number) %>% 
  mutate(sum_DCORR = sum(DCORR, na.rm = TRUE)) %>% 
  mutate(count = n())


# calculate average after group_by for each groups of diagnoses + visit number
temp2 <- temp2 %>% 
  group_by(diagnosis.f, visit_number) %>%
  mutate(avg_DCORR = sum_DCORR/count) %>% 
  mutate(se_DCORR = plotrix::std.error(DCORR)) 


ggplot(temp2, aes(x = visit_number, y = avg_DCORR, color = diagnosis.f)) + 
  geom_line(size = 0.7) + ggtitle("Time series of DCORR score") +
  scale_x_continuous(breaks=c(1, 2, 3, 4, 5, 6, 7)) + scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  geom_errorbar(aes(ymin = avg_DCORR - se_DCORR, ymax = avg_DCORR + se_DCORR), width = 0.2)

# bar graphs of visit number
ggplot(temp2, aes(x = visit_number)) + geom_bar() + ggtitle("DCORR visits distribution") +
  scale_x_continuous(breaks=c(1, 2, 3, 4, 5, 6, 7)) + 
  geom_text(stat='count', aes(label = ..count..), vjust = -0.2)
#ggsave("DCORR_visits.png", height = 7, width = 12)
  
  
#ggsave("DCORR_over_time.png", height = 7, width = 12)
# http://www.sthda.com/english/articles/32-r-graphics-essentials/128-plot-time-series-data-using-ggplot/
  
```


Add error bars to time series
Bar graphs of how many people completed which visit


### Now for all the other measures other than DCORR - DO NOT RUN IF NOT PLOTTING 
```{r}
### BNTCORR
bntcorr <- lang_behav_measures %>% 
  select(LONI_ID, diagnosis.f, visit_number, BNTCORR) %>% 
  filter(!is.na(BNTCORR))


# if we were to use visit numbers
# calculate sum of DCORR score based on groups of diagnoses type and visit number
bntcorr <- bntcorr %>% 
  group_by(diagnosis.f, visit_number) %>% 
  mutate(sum_BNTCORR = sum(BNTCORR, na.rm = TRUE)) %>% 
  mutate(count_BNTCORR = n()) %>% 
  mutate(avg_BNTCORR = sum_BNTCORR/count_BNTCORR, se_BNTCORR = plotrix::std.error(BNTCORR))

# time series of measure change
ggplot(bntcorr, aes(x = visit_number, y = avg_BNTCORR, color = diagnosis.f)) + 
  geom_line(size = 0.7) + ggtitle("Time series of BNTCORR score") +
  scale_x_continuous(breaks=c(1, 2, 3, 4, 5, 6, 7)) + scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  geom_errorbar(aes(ymin = avg_BNTCORR - se_BNTCORR, ymax = avg_BNTCORR + se_BNTCORR), width = 0.2)
#ggsave("BNTCORR_over_time.png", height = 7, width = 12)

# bar graphs of visit number
ggplot(bntcorr, aes(x = visit_number)) + geom_bar() + ggtitle("BNTCORR visits distribution") +
  scale_x_continuous(breaks=c(1, 2, 3, 4, 5, 6, 7)) + 
  geom_text(stat='count', aes(label = ..count..), vjust = -0.2)
#ggsave("BNTCORR_visits.png", height = 7, width = 12)


### ANCORR
ancorr <- lang_behav_measures %>% 
  select(LONI_ID, diagnosis.f, visit_number, ANCORR) %>% 
  filter(!is.na(ANCORR))


# if we were to use visit numbers
# calculate sum of DCORR score based on groups of diagnoses type and visit number
ancorr <- ancorr %>% 
  group_by(diagnosis.f, visit_number) %>% 
  mutate(sum_ANCORR = sum(ANCORR, na.rm = TRUE)) %>% 
  mutate(count_ANCORR = n()) %>% 
  mutate(avg_ANCORR = sum_ANCORR/count_ANCORR, se_ANCORR = plotrix::std.error(ANCORR))

ggplot(ancorr, aes(x = visit_number, y = avg_ANCORR, color = diagnosis.f)) + 
  geom_line(size = 0.7) + ggtitle("Time series of ANCORR score") +
  scale_x_continuous(breaks=c(1, 2, 3, 4, 5, 6, 7)) + scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  geom_errorbar(aes(ymin = avg_ANCORR - se_ANCORR, ymax = avg_ANCORR+ se_ANCORR), width = 0.2)
#ggsave("ANCORR_over_time.png", height = 7, width = 12)

# bar graphs of visit number
ggplot(ancorr, aes(x = visit_number)) + geom_bar() + ggtitle("ANCORR visits distribution") +
  scale_x_continuous(breaks=c(1, 2, 3, 4, 5, 6, 7)) + 
  geom_text(stat='count', aes(label = ..count..), vjust = -0.2)
#ggsave("ANCORR_visits.png", height = 7, width = 12)


### PPTP_COR
pptp <- lang_behav_measures %>% 
  select(LONI_ID, diagnosis.f, visit_number, PPTP_COR) %>% 
  filter(!is.na(PPTP_COR))


# if we were to use visit numbers
# calculate sum of DCORR score based on groups of diagnoses type and visit number
pptp <- pptp %>% 
  group_by(diagnosis.f, visit_number) %>% 
  mutate(sum_PPTPCORR = sum(PPTP_COR, na.rm = TRUE)) %>% 
  mutate(count_PPTPCORR = n()) %>% 
  mutate(avg_PPTPCORR = sum_PPTPCORR/count_PPTPCORR, se_PPTPCORR = plotrix::std.error(PPTP_COR))

ggplot(pptp, aes(x = visit_number, y = avg_PPTPCORR, color = diagnosis.f)) + 
  geom_line(size = 0.7) + ggtitle("Time series of PPTP_COR score") +
  scale_x_continuous(breaks=c(1, 2, 3, 4, 5, 6, 7)) + scale_y_continuous(breaks = scales::pretty_breaks(n = 10)) +
  geom_errorbar(aes(ymin = avg_PPTPCORR - se_PPTPCORR, ymax = avg_PPTPCORR + se_PPTPCORR), width = 0.2)
#ggsave("PPTP_COR_over_time.png", height = 7, width = 12)
# to see why PPTP plots look weird
weird <- pptp %>% 
  filter(diagnosis.f == "CON" & visit_number == 4 |
           diagnosis.f == "CON" & visit_number == 2 |
           diagnosis.f == "CON" & visit_number == 3) 

# bar graphs of visit number
ggplot(pptp, aes(x = visit_number)) + geom_bar() + ggtitle("PPTP_COR visits distribution") +
  scale_x_continuous(breaks=c(1, 2, 3, 4, 5, 6, 7)) + 
  geom_text(stat='count', aes(label = ..count..), vjust = -0.2)
#ggsave("PPTP_COR_visits.png", height = 7, width = 12)

```








```{r}

temp1 <- lang_behav_measures %>% 
  select(LONI_ID, diagnosis.f, visit_number, visits_time_diff, DCORR)
tt <- temp1 %>% 
  filter(diagnosis.f == "CON" & visit_number == 2) 

tt %>% 
  group_by()
  summarize(sum(DCORR))
#1891/126
  
  tt <- temp1 %>% 
  filter(diagnosis.f == "CON" & visit_number == 4) %>% 
  mutate(count = n())

# con2, 126

tt %>% 
  summarize(n())




# calculate sum of DCORR score based on groups of diagnoses type and time differences round
temp2 <- temp1 %>% 
  group_by(diagnosis.f, time_diff_round) %>% 
  summarize(sum_DCORR = sum(DCORR, na.rm = TRUE)) %>% 
  mutate(count = n())


```

